A new method to estimate dark matter halo concentrations

### 2015-01-21 paper/article.tex
finished section on new method description

% definition to produce a "greater than or similar to" symbol

@@

\title[New approach to Halo concentrations]{A new method to estimate
  dark matter halo concentrations}
\author[Poveda \& Forero-Romero]{
\parbox[t]{\textwidth}{\raggedright
  Christian Poveda$^{1}$ \&
  Jaime E. Forero-Romero$^{1}$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\
}
\maketitle

@@

\section{Introduction}
\label{sec:introduction}
In the concordance cosmology paradigm the the matter content of the
Universe are dominated by dark matter which behaves as a collisionless
fluid under the influence of gravity. In the last three decades
numerical experiments have made possible the simulation of large
volumes of a Universe dominated by dark matter.

One of the most striking results of these simulations is that dark
matter clumps on galactic scales follows a universal density profile
which in a first approximation is spherical symmetric and only
dependent on the radial coordinate. These profiles seem to be
universal; independent of the cosmological parameters and self-similar
for different spatial scapes after an addecuate re-scaling is
appplied.

@@

Once the density profile is expressed in dimensionless variables the
only free variable is the concentration.
There are two main methods to estimate this values in dark matter
halos extracted from N-body simulations.
The first method bins the particles in the logarithm of the radial
coordinate from the halo center to estimate the density in each
logarithmic bin.
Then a direct fit to the density as a function of the radial
coordinate is performed. This method has been most recently advocated
by \cite{Ludlow2014} to study the mass-concentration-redshift relation
of dark matter halos using the Millennium Simulation Series.

A second method uses the circular velocity profile.
As it was shown in Fig. (\ref{fig:velocity}) the circular velocity
shows a maximum for all profiles with concentration values larger than
$c>???$.
For the numerical halos wne can then find the value of $x$ for which
the normalized circular velocity $v(<x)$ shows a maximum and solve
numerically for the corresponding value of the concentration using
Eq. (\ref{eq:max_v}).  This method has been used by \cite{Klypin2014}
to study the mass-concentration-redshift relation using the Multidark
Simulation Suite.

We estimate the concentration parameter by fitting the
integrated mass profile described in Equation
(\ref{eq:profile}). First we define the center of the halo to be  at
the position of the particle with the lowest gravitational
potential.

We construct the integrated mass profile by ranking the particles by
their increasing distance to the center of the halo. Once they are ranked,
the total mass at a radius $r_i$, increases by $m_p$, where $r_i$ is
the position of the $i$-th particle and $m_p$ is the mass of the
computational particle.  In this process we discard the
particle at the center.

We stop the construction of the integrated mass profile once we arrive
at an average density of $\Delta_h\bar{\rho}$, with $\Delta_h=740$,
roughly corresponding to 200 times the critical density. This radius marks the
virial radius and the virial mass. We divide the total mass enclosed
mass $M_i$ and the radii $r_i$ by these values to obtain the
dimensionless variables $x_i$ and $m_i$.

Using these new variables we define the following $\chi^2$ function

\begin{equation}
\chi^2(c) = \sum_{i}[\log m_i - \log m(< x_i;c)]^2,
\end{equation}
%
where $m(<x_i;c)$ corresponds to the values in Eq.(\ref{eq:profile}) at
$x=x_i$ for a given value of the concentration parameter $c$.

We sample the likelihood function distribution defined by ${\cal
  L}(c)=\exp(-\chi^2(c)/2)$ with a Metropolis-Hastings algorithm to
find the optimum value of $c$ and its  associated uncertainty
$\sigma_c$. We used 50000 steps for walk, where each step was randomly generated
with a normal distribution centered around the last step with an standard
deviation of $\sigma=0.03$.

>>>>

@@

\title[A new method for concentrations]{A new method to estimate
  dark matter halo concentrations}
\author[Poveda \& Forero-Romero]{
\parbox[t]{\textwidth}{\raggedright
  Christian Poveda$^{1}$ \&
  Jaime E. Forero-Romero$^{1}$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia\\
}
\maketitle

@@

\section{Introduction}
\label{sec:introduction}
In the concordance cosmology paradigm the the matter content of the
Universe is dominated by dark matter which behaves as a collisionless
fluid under the influence of gravity. In the last three decades
numerical experiments have made possible the simulation of dark matter
dominated universes, yielding valuable insights on the large scale
structure formation process as depicted in observations.

One of the most striking results of these simulations is that dark
matter clumps on galactic scales follow a universal density profile
which in a first approximation is spherical symmetric and only
dependent on the radial coordinate. These profiles seem to be
universal; independent of the cosmological parameters and self-similar
for different spatial scapes after an addecuate re-scaling is
appplied.

@@

As we saw in the previous sections, once the density profile is
expressed in dimensionless variables the only free parameter in the
density profile  is the concentration. There are two main methods to
estimate concentrations in dark matter halos extracted from N-body
simulations.

The first method tries to directly estimate the density profile. 
It takes all the particles in the halo and bins them in the logarithm
of the radial coordinate from the halo center.
Then, it estimates the density in each logarithmic bin counting the
particles and divinding by the corresponding shell volume.
At this point is is possible to make a direct fit to the density as a
function of the radial coordinate. 
This method has been most recently used by \cite{Ludlow2014} to study
the mass-concentration-redshift relation of dark matter halos using
the Millennium Simulation Series.

A second method uses the circular velocity profile.
As it was shown in Fig. (\ref{fig:velocity}) the circular velocity
shows a maximum for all profiles with concentration values larger than
$c>2$.
The method finds the value of $x$ for which the normalized circular
velocity $v(<x)$ shows a maximum. 
Using this value it solves numerically for the corresponding value of
the concentration using Eq. (\ref{eq:max_v}).  
This method has been most recently used by \cite{Klypin2014} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

Our method is a third option that uses the integrated mass profile.
First we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.
Then we rank the particles by their increasing radial distance from
the center.
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is
the position of the $i$-th particle and $m_p$ is the mass of a single
computational particle.
In this process we discard the particle at the center.

We stop the construction of the integrated mass profile once we arrive
at an average density of $\Delta_h\bar{\rho}$, with $\Delta_h=740$,
roughly corresponding to 200 times the critical density. 
This radius marks the virial radius and the virial mass. 
We divide the enclosed mass mass $M_i$ and the radii $r_i$ by these
virial values to obtain the dimensionless variables $m_i$ and $x_i$.

The construction of the numerical integrated mass profile has the
advantage that it does not involve any binning and uses the
information from all the particles in the halo, unlike the method that
tries to directly build. 
Furthermore, as it will be clear in the next paragraph, the fit of
this computational profile to the analytic expectation uses the
information from all points, not only a single maxima point as the
method using the circular velocity profile.

We use a Metropolis-Hastings algorithm to sample the likelihood
function distribution defined by ${\cal L}(c)=\exp(-\chi^2(c)/2)$
where the $\chi^2(c)$ is written as

\begin{equation}
\chi^2(c)= \sum_{i=1}^{N}[\log m_i - \log m(< x_i;c)]^2,
\end{equation}
%
where $m(<x_i;c)$ corresponds to the values in Eq.(\ref{eq:profile}) at
$x=x_i$ for a given value of the concentration parameter $c$ and the
$i$ index sums over all the particles in the numerical profile.

For the walk in the Metropolis-Hastings algorithm we used 50000 steps
where each step was randomly generated with a normal distribution
centered around the last step with an standard deviation of
$\sigma=0.03$. 
From the $\chi^2$ distribution we find the optimal value of the
concentration and  its associated uncertainty.
